Overlapping Batches for the Assessment of Solution Quality in Stochastic Programs

### 2013-10-26 omrp.tex
Streamlining Sec 5 - II
A whole bunch of edits

In the previous section, we proved asymptotic results regarding the OMRP estimators. 
In this section, we apply OMRP to several test problems and examine its small-sample behavior. 
We begin our discussion by introducing the test problems used for numerically evaluating OMRP in \S \ref{ssec:test}, followed by experimental setup in \S \ref{ssec:expS}. Then, we present and discuss our numerical experiments in \S \ref{ssec:results}.

@@

We empirically tested the effectiveness of OMRP on four test problems.
All test problems are two-stage stochastic programs with recourse and were selected to study the behavior of OMRP under a variety of different situations, as we will shortly discuss below.
The characteristics of these problems are given in Table \ref{tb:test_problems}. 
The first test problem is a newsvendor problem with demand distributed uniformly on $\{0, 0.0001, 0.0002, \dots, 10\}$.
We denote this problem as NVD.
The instance of the newsvendor problem can be found in \citep{Bayraksan2006}, except here we used a discrete uniform demand distribution.
The second is a capacity expansion planning problem, denoted CEP1, with random demand vector $\xit$ of dimension 3, with 216 total realizations \citep{higle1996}.
The third is a stochastic knapsack problem denoted 10D, with 10 stochastic parameters and 10 binary decision variables \citep{kleywegt2002sample}.
The final problem is a stochastic vehicle routing problem denoted DB1, with 46 stochastic parameters and a total of $4.5 \cdot 10^{25}$ realizations \citep{donohue1995upper}.

@@

The instance of the newsvendor problem as well as the candidate solutions used for both problems can be found in \citep{Bayraksan2006}, except here we used a discrete uniform demand distribution for the newsvendor problem.

We made several considerations in selecting the test problems.
Newsvendor can be solved very quickly, which allows for solving a large number of independent runs for examining the small sample behavior of OMRP.
CEP1 is also easy to solve, and obtains the optimal solution $\xs$ with high probability even for small sample sizes.
This feature shows the behavior $VG$ when the gap size is determined mostly by sampling error, rather than bias.
10D demonstrates the behavior of OMRP when $|X|$ is finite.
Furthermore, 10D has a high condition number \citep{kleywegt2002sample}, making it unlikely that the optimal solution $\xs$ is found when using a sampling solution, thus demonstrating the behavior of OMRP with the bias of the gap is significant.
The optimal solutions of both 10D and CEP1 are known, allowing for explicitly checking the computational results.
In contrast, DB1 is a large scale example, with an unknown optimal solution.
This final problems allows us to show that OMRP works on more computationally difficult problems.

@@

We used a total sample size of $n = 30 m$, with several values of the batch size $m=100, 200, \ldots$. 
Then, at each value of $m$, we formed OMRP estimators, with varying amounts of overlap. 
We also formed 90\% confidence intervals on the optimality gap using ORMP estimators. 
Results for the newsvendor problem were compiled over 10,000 independent runs, while the CEP1 and 10D results were taken over 1,000 independent runs, and DB1 results were generated over 500 independent runs.

While NVD can be written as a two-stage stochastic linear program with recourse, it can be solved quickly by sorting the demand values. 
We solved it using the quicksort algorithm in C++. 
The problems CEP1 and DB1 are two-stage stochastic linear programs with recourse and were solved with the regularized decomposition algorithm of \citep{ruszczynski1986regularized}, using the accelerated version of the algorithm implemented in C++ \citep{ruszczynski1997accelerating}. 
10D is a stochastic integer program and was solved using dynamic programming.  
In all cases, we used the Mersenne Twister algorithm to generate random samples \citep{wagner_twister_09}.

@@

Summary of results from these experiments are depicted in Figures~\ref{fig:nv} and \ref{fig:varvar}.

\begin{figure}[htb!]
	\centering
	\subfloat[][Variance Reduction]{
	   \label{fig:nv_varvar}
	   \includegraphics[width=.32\textwidth]{images/nv_varvar_discrete}}
	\subfloat[][Expected Variance]{
	   \label{fig:nv_avgvar}
	   \includegraphics[width=.32\textwidth]{images/nv_avgvar_discrete}}
	\subfloat[][Coverage Probability]{
	   \label{fig:nv_cover}
	   \includegraphics[width=.32\textwidth]{images/nv_cover_discrete}}
	\caption{
		Summary of results for the newsvendor problem. 
		(a) Reduction in variance of $nVG$,
		(b) estimates of $\e{nVG}$, and 
		(c) coverage probability of the confidence intervals for various values of $\gamma/m$ ($\gamma/m=1$ denotes the nonoverlapping batches).
	}
\label{fig:nv}
\end{figure}

@@

Figures~\ref{fig:nv}(a) and \ref{fig:varvar} show the reduction of variance, with each term being normalized with respect to $\var{nVG}$ of the nonoverlapping case (MRP). 
The solid lines in these figures show the theoretical variance reduction from \citep{Welch1987} given by (\ref{eq:var_reduct_formula}). 
Empirical results agree with the theoretical values well, with more variability observed in CEP1 and 10D, providing evidence that similar variance reduction can be achieved in stochastic programming by overlapping the batches even with small sample sizes.
DB1 shows some variance reduction, but less than the asymptotic amount for small sample sizes.

Figure~\ref{fig:nv}(b) shows estimates of $\e{nVG}$ in the newsvendor problem for changing batch size and amount of overlap. 
We can see that the expectation of the variance estimator is not changed with increasing overlap. 
Although the estimate of $\e{nVG}$ looks large for $m=100$, the error from $\sigma^2_\xh$ is never more than 1\%.  
Estimates for $\e{nVG}$ showed a similar pattern for CEP1 and 10D.  Those graphs are omitted for brevity.

Finally, Figure~\ref{fig:nv}(c) shows the coverage probability of confidence intervals generated by OMRP for several values of $m$ across varying values of $\gamma/m$ for the newsvendor problem.  
The results of the classical overlapping batches estimators show that coverage probability does not change with the amount of overlap, which we can  empirically see in this figure for OMRP for newsvendor, and we observed the same for CEP1 and 10D.  
The coverage probabilities from applying the (nonoverlapping) MRP algorithm to these problems presented in \citep{Bayraksan2006} agree with the our results: for the newsvendor problem, coverage probability drops as $m$ increases and the coverage probability of CEP1 (not shown) remains fairly constant around the desired value of $90\%$. 
Coverage probability for 10D and DB1 (not shown) were very high, more than $0.99$.  The bias from solving (\ref{eq:sto_prog_m}) for these larger problems seems to be much more significant than the variance.

Because variance reduction drops quickly as $\gammab$ drops below $1$, the authors recommend using an intermediate value of $\gammab$, such as $1/3$ or $1/4$, to gain the majority of the variance reduction, while reducing the number of optimization problems to be solved. 
Warm-starting the algorithm to solve the sampling problems, when such as scheme is available, can significantly reduce the solution times.

\section{Conclusion} \label{sec:concl}
We have extended previous work from simulation analysis to assessing solution quality in stochastic programming by means of variably overlapping batches \citep{Meketon1984,Song1992,Welch1987} for Monte Carlo sampling-based estimators of optimality gaps in stochastic programs \citep{Mak1999}. 
We have provided conditions under which the resulting point estimator of the optimality gap and its associated sample variance are consistent, the same variance reduction can be achieved, and asymptotically valid confidence intervals can be formed. 
Empirical results with small sample sizes indicate that the asymptotic reductions in variance of the variance estimator show a similar decrease in OMRP, while bias and coverage probability remain unaffected.

>>>>

In the previous section, we proved asymptotic results regarding the OMRP estimators. 
In this section, we apply OMRP to several test problems and examine its small-sample behavior. 
We begin our discussion by introducing the test problems used for numerically evaluating OMRP in \S \ref{ssec:test}, followed by our experimental setup in \S \ref{ssec:expS}. Then, we present and discuss our numerical experiments in \S \ref{ssec:results}.

@@

We empirically tested the effectiveness of OMRP on four test problems.
All test problems are two-stage stochastic programs with recourse and were selected to study the behavior of OMRP under a variety of different situations---we will shortly discuss this below.
The characteristics of these problems are given in Table \ref{tb:test_problems}. 
The first test problem is a newsvendor problem with demand distributed uniformly on $\{0, 0.0001, 0.0002, \dots, 10\}$.
We denote this problem as NVD.
The instance of the newsvendor problem can be found in \citep{Bayraksan2006}, except here we used a discrete uniform demand distribution.
The second is a capacity expansion planning problem, denoted CEP1, with random demand vector $\xit$ of dimension 3, with 216 total realizations \citep{higle1996}.
The third is a stochastic knapsack problem denoted 10D, with 10 stochastic parameters and 10 binary decision variables \citep{kleywegt2002sample}.
The final problem is a stochastic vehicle routing problem, denoted DB1, with 46 stochastic parameters and a total of $4.5 \times 10^{25}$ realizations \citep{donohue1995upper}.

@@

We made several considerations in selecting the test problems.
NVD can be solved very quickly, which allows for solving a large number of independent runs for examining the small-sample behavior of OMRP.
CEP1 is also easy to solve; the optimal solution $\xs$ can be obtained with high probability even for small sample sizes.
This feature shows the behavior of $VG$ when the gap size is determined mostly by sampling error, rather than bias.
10D demonstrates the behavior of OMRP when $|X|$ is finite.
Furthermore, 10D has a high condition number \citep{kleywegt2002sample}, making it unlikely that the optimal solution $\xs$ is found when using a sampling solution, thus demonstrating the behavior of OMRP when the bias of the gap is significant.
The optimal solutions of NVD, CEP1, and 10D are known, allowing for explicitly checking and verifying the results.
In contrast, DB1 is a large-scale problem with an unknown optimal solution.
This final test problem allows us to investigate how OMRP works on large-scale problems.

The candidate solutions used for NVD and CEP1 and their characteristics can be found in \citep{Bayraksan2006}.
We used the candidate solution (0, 0, 1, 1, 1, 0, 1, 1, 0, 1) for 10D with $\mu_{\xh} = 158.579$ and the candidate solution (11, 14, 8, 11, 7) with an estimated $\mu_{\xh} = 1.08$ for DB1. 
Because the optimal solution of DB1 is unknown, a Latin hypercube sample of size 50,000 was used to estimate $\mu_{\xh}$, which was then used to estimate the coverage probability of the OMRP CIs for this test problem below.

@@

Our experiments were conducted as follows. 
We used a total sample size of $n = 30 m$, with several values of the batch size $m=100, 200, \ldots$. 
Then, at each value of $m$, we formed the OMRP estimators with varying amounts of overlap. 
We also set $\alpha=0.10$ formed 90\% confidence intervals on the optimality gap using the ORMP estimators. 
Results for NVD were compiled over 10,000 independent runs, while the CEP1 and 10D results were taken over 1,000 independent runs, and DB1 results were generated over 500 independent runs.

We solved the sampling approximations of the test problems as follows.
While NVD can be written as a two-stage stochastic linear program with recourse, it can be solved quickly by sorting the demand values. 
We solved it using the quicksort algorithm in C++. 
The problems CEP1 and DB1 are two-stage stochastic linear programs with recourse and were solved with the regularized decomposition algorithm of \citet{ruszczynski1986regularized}, using the accelerated version of the algorithm implemented in C++ \citep{ruszczynski1997accelerating}. 
10D is a stochastic knapsack problem and was solved efficiently using dynamic programming.  
In all cases, we used the Mersenne Twister algorithm to generate random samples \citep{wagner_twister_09}.

@@

Summary of results from these experiments are depicted in Figures~\ref{fig:nv} and \ref{fig:varvar}.
Figures~\ref{fig:nv}(a) and \ref{fig:varvar} show the reduction of variance, with each term being normalized with respect to $\var{nVG}$ of the nonoverlapping case (MRP). 
The solid lines in these figures show the theoretical variance reduction from \citep{Welch1987} given by (\ref{eq:var_reduct_formula}). 
Empirical results agree with the theoretical values well, with more variability observed in CEP1, 10D, and DB1. 
These experiments suggest that similar variance reduction can be achieved in stochastic programming by overlapping the batches, even with small sample sizes.
We note that the variance reduction in DB1 can be less than the asymptotic amount for the small sample sizes used in these experiments. 
Even with the relatively small sample sizes, compared to the size of this problem, overlapping appears to achieve some variance reduction.

\begin{figure}[htb!]
	\centering
	\subfloat[][Variance Reduction]{
	   \label{fig:nv_varvar}
	   \includegraphics[width=.32\textwidth]{images/nv_varvar_discrete}}
	\subfloat[][Expected Variance]{
	   \label{fig:nv_avgvar}
	   \includegraphics[width=.32\textwidth]{images/nv_avgvar_discrete}}
	\subfloat[][Coverage Probability]{
	   \label{fig:nv_cover}
	   \includegraphics[width=.32\textwidth]{images/nv_cover_discrete}}
	\caption{
		Summary of results for NVD. 
		(a) Reduction in variance of $nVG$,
		(b) estimates of $\e{nVG}$, and 
		(c) coverage probability of the confidence intervals for various values of $\gamma/m$ ($\gamma/m=1$ denotes the nonoverlapping batches).
	}
\label{fig:nv}
\end{figure}

@@

Figure~\ref{fig:nv}(b) shows estimates of $\e{nVG}$ in the newsvendor problem for changing batch size and amount of overlap. 
We can see that the expectation of the variance estimator is not changed with increasing overlap. 
Although the estimate of $\e{nVG}$ looks large for $m=100$, the error from $\sigma^2_\xh$ is never more than 1\%.  
Estimates for $\e{nVG}$ showed a similar pattern for CEP1, 10D, and DB1.  
Those graphs are omitted for brevity.

Finally, Figure~\ref{fig:nv}(c) shows the coverage probability of confidence intervals generated by OMRP for several values of $m$ across varying values of $\gamma/m$ for the newsvendor problem.  
The results of the classical overlapping batches estimators show that coverage probability does not change with the amount of overlap, which we can  empirically see in this figure for OMRP for NVD.
We observed the same for CEP1, 10D, and DB1.  
The coverage probabilities from applying the (nonoverlapping) MRP algorithm to these problems presented in \citep{Bayraksan2006} agree with the our results: for the newsvendor problem, coverage probability drops as $m$ increases and the coverage probability of CEP1 (not shown) remains fairly constant around the desired value of $90\%$. 
Coverage probability for 10D and DB1 (not shown) were very high, more than $0.99$.  
The bias from solving (\ref{eq:sto_prog_m}) for these problems seems to be much more significant than the variance.

Because variance reduction drops quickly as $\gammab$ drops below $1$, the authors recommend using an intermediate value of $\gammab$, such as $1/3$ or $1/4$, to gain the majority of the variance reduction, while reducing the number of optimization problems to be solved. 
Warm-starting the algorithm to solve the sampling problems, when such as scheme is available, can significantly reduce the computation times.

\section{Conclusions} \label{sec:concl}
We have extended previous work from simulation analysis to assessing solution quality in stochastic programming by means of variably overlapping batches \citep{Meketon1984,Song1992,Welch1987} for Monte Carlo sampling-based estimators of optimality gaps in stochastic programs \citep{Mak1999}. 
We have provided conditions under which the resulting point estimator of the optimality gap and its associated sample variance are consistent, the same variance reduction can be achieved, and asymptotically valid confidence intervals can be formed. 
Empirical results with small sample sizes indicate that the asymptotic reductions in variance of the variance estimator show a similar decrease in OMRP, while bias and coverage probability remain unaffected.
